Ab Initio Effective One-Electron Potentials Revisited: General Theory and Example Applications for Charge-Transfer Energy Calculations

### 2019-07-10 papers/2_oep_technique/paper-r0.tex
Paper revision and basis set module minirevision

\author{Marta Cho{\l}uj} 
\author{Joanna Bednarska}
\author{Wojciech Bartkowiak}

@@

The one\hyp{}electron potential $v^{\rm eff}({\bf r})$
produced by a certain effective one\hyp{}electron charge density distribution $\rho^{\rm eff}({\bf r})$
is 
%
\begin{equation} \label{e:v-eff}
	v^{\rm eff}({\bf r}) = \int \frac{ \rho^{\rm eff}({\bf r}') }{ \vert {\bf r}' - {\bf r} \vert} d{\bf r}' \;,
\end{equation}
%
where ${\bf r}$ is a spatial coordinate. 
For convenience, the total effective density can be split into nuclear and electronic contributions,
%
\begin{equation} \label{e:rho-eff}
 \rho^{\rm eff}({\bf r}) = \lambda \rho^{\rm eff}_{\rm nuc}({\bf r}) + \rho^{\rm eff}_{\rm el}({\bf r}) \;,
\end{equation}
%
where $\lambda$ is a certain parameter and is assumed either 1 or 0 in this work.
Formally, the electronic part can be expanded in terms of an effective
one\hyp{}particle density matrix represented in a certain basis of orbitals, $\BM{\phi}({\bf r})$,
%
\begin{equation} \label{e:d-spectral}
	\rho^{\rm eff}_{\rm el}({\bf r}) = \sum_{\alpha\beta} D_{\alpha\beta}^{\rm eff} 
	\phi_\alpha({\bf r}) \phi_\beta^{*}({\bf r})  \;.
\end{equation}
%
Based on that, the operator form of the effective potential 
can be written as
%
\begin{equation} \label{e:oep-operator}
	\hat{v}^{\rm eff} = 
        \lambda \hat{v}_{\rm nuc} +
        \sum_{\alpha\beta} 
	\Ket{\alpha} 
	\left[
	\int
	\hat{J}^{\rm}_{\alpha\beta}({\bf r}) 
	\rho^{\rm eff}({\bf r})
	d{\bf r} 
	\right]
	\Bra{\beta}  
\end{equation}
%
with the nuclear
and Coulomb operators defined by
%
\begin{subequations}
\begin{align}
        \hat{v}_{\rm nuc} &\equiv \\
	\hat{J}^{\rm}_{\alpha\beta}({\bf r}) f({\bf r}) &\equiv
	f({\bf r}) 
	\int
	\frac{ \phi_\alpha^{*}({\bf r}') \phi_\beta({\bf r}') }{ \vert {\bf r}' - {\bf r} \vert} d{\bf r}'
\end{align}
\end{subequations}
%
for any one\hyp{}electron function $f({\bf r})$. The matrix element of the effective potential operator
is therefore given by
%
\begin{equation}
	\tBraKet{\alpha}{ \hat{v}^{\rm eff} }{\beta}
	= \lambda \sum_x^{\rm At} W_{\alpha\beta}^{(x)} +
        \sum_{\gamma\delta} \BraKet{\alpha\beta}{\gamma\delta} D^{\rm eff}_{\gamma\delta}  \;,
\end{equation}
%
where 
%
\begin{equation}
 W_{\alpha\beta}^{(x)} = 
 Z_x \int \frac{\phi^*_\alpha({\bf r}) \phi_\beta({\bf r})}{\vert {\bf r} - {\bf r}_x \vert} d{\bf r} \;,
\end{equation}
%
$Z_x$ is the atomic number of the $x$th atom,
and the symbol `At' denotes all atoms that contribute to the effective potential.
In the above equation and throughout the work, 
%
\begin{equation}
\tBraKet{\alpha}{\mathcal{O}(1)}{\beta} \equiv \int \phi^*_\alpha({\bf r}) \mathcal{O}(1) \phi_\beta({\bf r}) d{\bf r} 
\end{equation}
%
for any one\hyp{}electron operator $\mathcal{O}(1)$ 
and the electron repulsion integral (ERI)
is defined according to
%
\begin{equation}
	\BraKet{\alpha\beta}{\gamma\delta} \equiv
	\iint 
	\frac{ \phi_\alpha^{*}({\bf r}_1) \phi_\beta({\bf r}_1) 
	       \phi_\gamma^{*}({\bf r}_2) \phi_\delta({\bf r}_2) }{ \vert {\bf r}_1 - {\bf r}_2 \vert}
	d{\bf r}_1 d{\bf r}_2  \;.
\end{equation}
%

@@

As a starting point we consider perturbation theory model of Murrell et al.,
in which repulsive energy is originally given by
%
\begin{equation}
 \Delta^{AB} E^{\rm Rep} = 
 \Delta^{AB} E^{\rm Rep}(\mathcal{O}(S)) + 
 \Delta^{AB} E^{\rm Rep}(\mathcal{O}(S^2))
\end{equation}
%
where the first\hyp{}order term with respect to the intermolecular overlap is
%
\begin{multline}
  \Delta^{AB}  E^{\rm Rep}(\mathcal{O}(S)) = -2\sum_{a\in A} \sum_{b\in B}
               S_{ab} \Big\{
           V^A_{ab} 
         + V^B_{ab} \\
 + \sum_{c\in A} \left[ 2(ab \vert cc) - (ac \vert bc) \right] 
 + \sum_{d\in B} \left[ 2(ab \vert dd) - (ad \vert bd) \right]
                \Big\}
\end{multline}
%
and the second\hyp{}order term is given in Eq.(XXX) in Ref. %cite
In this work, we focus to simplify only the former term because this is usually
the bottleneck of evaluating the Pauli repulsive contributions to
forces in molecular dynamics when using the effective potential method, whereas
very inexpensive approximation to the latter term has already been worked out previously. %cite

>>>>

\author{Marta Cho{\l}uj} 
\author{Joanna D. Bednarska}
\author{Wojciech Bartkowiak}

@@

The one\hyp{}electron potential $v^{\rm eff}({\bf r})$
produced by a certain effective one\hyp{}electron charge density distribution $\rho^{\rm eff}({\bf r})$
is 
%
\begin{equation} \label{e:v-eff}
	v^{\rm eff}({\bf r}) = \int \frac{ \rho^{\rm eff}({\bf r}') }{ \vert {\bf r}' - {\bf r} \vert} d{\bf r}' \;,
\end{equation}
%
where ${\bf r}$ is a spatial coordinate. 
For convenience, the total effective density can be split into nuclear and electronic contributions,
%
\begin{equation} \label{e:rho-eff}
 \rho^{\rm eff}({\bf r}) = \lambda \rho^{\rm eff}_{\rm nuc}({\bf r}) + \rho^{\rm eff}_{\rm el}({\bf r}) \;,
\end{equation}
%
where $\lambda$ is a certain parameter and is assumed either 1 or 0 in this work.
Formally, the electronic part can be expanded in terms of an effective
one\hyp{}particle density matrix represented in a certain basis of orbitals, $\BM{\phi}({\bf r})$,
%
\begin{equation} \label{e:d-spectral}
	\rho^{\rm eff}_{\rm el}({\bf r}) = \sum_{\alpha\beta} D_{\alpha\beta}^{\rm eff} 
	\phi_\alpha({\bf r}) \phi_\beta^{*}({\bf r})  \;.
\end{equation}
%
Based on that, the operator form of the effective potential 
can be written as
%
\begin{equation} \label{e:oep-operator}
	\hat{v}^{\rm eff} = 
        \lambda \hat{v}_{\rm nuc} +
        %\sum_{\alpha\beta} 
	%\Ket{\alpha} 
        \int d{\bf r} \Ket{{\bf r}} 
	\left[
	\int
	\hat{J}^{\rm}_{\alpha\beta}({\bf r}') 
	\rho^{\rm eff}({\bf r}')
	d{\bf r}' 
	\right]
	%\Bra{\beta}  
        \Bra{{\bf r}}
\end{equation}
%
with the nuclear
and Coulomb operators defined by
%
\begin{subequations}
\begin{align}
        \hat{v}_{\rm nuc} &\equiv \\
	\hat{J}^{\rm}_{\alpha\beta}({\bf r}) f({\bf r}) &\equiv
	f({\bf r}) 
	\int
	\frac{ \phi_\alpha^{*}({\bf r}') \phi_\beta({\bf r}') }{ \vert {\bf r}' - {\bf r} \vert} d{\bf r}'
\end{align}
\end{subequations}
%
for any one\hyp{}electron function $f({\bf r})$. The matrix element of the effective potential operator
is therefore given by
%
\begin{equation}
	\tBraKet{\alpha}{ \hat{v}^{\rm eff} }{\beta}
	= \lambda \sum_x^{\rm At} W_{\alpha\beta}^{(x)} +
        \sum_{\gamma\delta} \BraKet{\alpha\beta}{\gamma\delta} D^{\rm eff}_{\gamma\delta}  \;,
\end{equation}
%
where 
%
\begin{equation}
 W_{\alpha\beta}^{(x)} = 
 Z_x \int \frac{\phi^*_\alpha({\bf r}) \phi_\beta({\bf r})}{\vert {\bf r} - {\bf r}_x \vert} d{\bf r} \;,
\end{equation}
%
$Z_x$ is the atomic number of the $x$th atom,
and the symbol `At' denotes all atoms that contribute to the effective potential.
In the above equation and throughout the work, 
%
\begin{equation}
\tBraKet{\alpha}{\mathcal{O}(1)}{\beta} \equiv \int \phi^*_\alpha({\bf r}) \mathcal{O}(1) \phi_\beta({\bf r}) d{\bf r} 
\end{equation}
%
for any one\hyp{}electron operator $\mathcal{O}(1)$ 
and the electron repulsion integral (ERI)
is defined according to
%
\begin{equation}
	\BraKet{\alpha\beta}{\gamma\delta} \equiv
	\iint 
	\frac{ \phi_\alpha^{*}({\bf r}_1) \phi_\beta({\bf r}_1) 
	       \phi_\gamma^{*}({\bf r}_2) \phi_\delta({\bf r}_2) }{ \vert {\bf r}_1 - {\bf r}_2 \vert}
	d{\bf r}_1 d{\bf r}_2  \;.
\end{equation}
%

@@

As a starting point we consider perturbation theory model of Murrell et al.,
in which repulsive energy is originally given by
%
\begin{equation}
 \Delta^{AB} E^{\rm Rep} = 
 \Delta^{AB} E^{\rm Rep}(S^{-1}) + 
 \Delta^{AB} E^{\rm Rep}(S^{-2}) \;,
\end{equation}
%
where the first\hyp{}order term with respect to the intermolecular overlap is
%
\begin{multline} \label{e:rep.murrell-etal.S1}
  \Delta^{AB}  E^{\rm Rep}(S^{-1}) = -2\sum_{i\in A}^{\rm Occ} \sum_{j\in B}^{\rm Occ}
               S_{ij} \Big\{
           V^A_{ij} 
         + V^B_{ji} \\
 + \sum_{k\in A}^{\rm Occ} \left[ 2\BraKet{ij}{kk} - \BraKet{ik}{jk} \right] 
 + \sum_{l\in B}^{\rm Occ} \left[ 2\BraKet{ij}{ll} - \BraKet{il}{jl} \right]
                \Big\}
\end{multline}
%
and the second\hyp{}order term is given in Eq.(XXX) in Ref. %cite
In this work, we focus to simplify only the former term because this is usually
the bottleneck of evaluating the Pauli repulsive contributions to
forces in molecular dynamics when using the effective potential method, whereas
very inexpensive approximation to the latter term has already been worked out previously. %cite
Note that the Coulomb and exchange integrals can be re\hyp{}cast as follows:
%
\begin{subequations}
 \begin{align}
 \BraKet{ij}{kk} &\equiv \sum_{\mu\in A} 
     C_{\mu i}^A \tBraKet{\mu}{\hat{v}_{kk}}{j} \;, \\
 \BraKet{ik}{jk} &\equiv \sum_{\mu\in A} 
     C_{\mu k}^A \tBraKet{\mu}{\hat{v}_{ik}}{j} \;.
 \end{align}
\end{subequations}
%
%Similarly, the exchange integrals reduce to
%
%\begin{equation}
% \BraKet{ik}{jk} \equiv \sum_{\mu\in A} 
%     C_{\mu k}^A \tBraKet{\mu}{\hat{v}_{ik}}{j} \;.
%\end{equation}
%
In the above equations, the auxiliary potential operators are given by
%
\begin{equation}
  \hat{v}_{ik} \equiv \int d{\bf r} \Ket{{\bf r}} 
        \left[
        \int d{\bf r}' \frac{\phi_i^{*}({\bf r}') \phi_k({\bf r}')}{\vert {\bf r}' - {\bf r}\vert}
        \right] \Bra{{\bf r}} \;.
\end{equation}
%
Now, using the prescription in Eq.~\eqref{e:ft-reduction} one can define a joint OEP
operator from nuclear, Coulomb and exchange parts as
%
\begin{equation}
 V^A_{ij} + 
 \sum_{k\in A}^{\rm Occ} 
  \left\{ 2\BraKet{ij}{kk} - \BraKet{ik}{jk} \right\}
\equiv \sum_{\mu\in A} \tBraKet{\mu}{ 
\hat{v}^{A[\mu i]}_{\rm eff}
 }{j}
\end{equation}
%
with
%
\begin{equation}
 \hat{v}^{A[\mu i]}_{\rm eff} \equiv C_{\mu i} \hat{v}^A_{\rm nuc} + 
 \sum_{k\in A}^{\rm Occ} \left[
 2C_{\mu i}^A \hat{v}^A_{kk} - C_{\mu k}^A \hat{v}^A_{ik}
 \right] \;.
\end{equation}
%
%The nuclear part can also be incorporated within the effective potential
%resulting in the following reduction
%
%
On the other hand, it immediately follows that
%
\begin{equation}
 \hat{v}^{A[\mu i]}_{\rm eff} \Ket{\mu} = 
  \sum_{k\in A}^{\rm Occ} \left\{
     2\hat{v}^A_{kk} \Ket{i} - \hat{v}^A_{ik} \Ket{k}
  \right\} \;.
\end{equation}
%
For practical calculations, the right hand side of the above equation can be expanded 
in the auxiliary basis,
%
\begin{equation}
  \sum_{k\in A}^{\rm Occ} \left\{
     2\hat{v}^A_{kk} \Ket{i} - \hat{v}^A_{ik} \Ket{k}
  \right\} \cong
  \sum_{\xi\in A}^{\rm DF} 
  V_{\xi i}^A \Ket{\xi} \;,
\end{equation}
%
where the matrix ${\bf V}^A$ can be considered as effective fragment parameters.
Doing the same operations on the twin operators associated with the molecule $B$
original theory of Murrell et. al from Eq.~\eqref{e:rep.murrell-etal.S1}
reduces to
%
\begin{equation}
  \Delta^{AB}  E^{\rm Rep}(S^{-1}) \cong 
 -2\sum_{i\in A}^{\rm Occ} \sum_{j\in B}^{\rm Occ}
               S_{ij} \Big\{
           \sum_{\xi \in A}^{\rm DF} V_{\xi i}^A S_{\xi j}
         + \sum_{\eta\in B}^{\rm DF} V_{\eta j}^B S_{\eta i}
                \Big\}
\end{equation}
%
which is a fully OEP\hyp{}based result. Note, that now only 
overlap integrals, which are relatively computationally very inexpensive,
need to be evaluated.
